### 2023-02-14 talk/talk.tex
various talk updates

Joint work with\\
Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King

@@

  1--5 are largely solved, from a methodological perspective.\\
  7 is exciting but not the focus of this talk.\\
  6 is our topic.

@@

\begin{frame}{Example: Pre-vaccination measles in England \& Wales}

@@

    Hence, likelihood-based methods.

      \item  In the time series case, iterated particle filtering (IF2) implemented in the R package \texttt{pomp} enables Masters-level statisticians to do this (\url{https://ionides.github.io/531w22/}). The science may be hard, but the statistics is becoming routine.
      \end{itemize}
  \end{frame}

@@

\item \myemph{Decoupling} leads to panel data analysis, by assumption. Iterated filtering methods extend to panel data (Breto et al, {\it Journal of the American Statistical Association}, 2019).

\item We must decide which parameters should be modeled as \myemph{shared} vs \myemph{unit-specific}.

\item The consequences of decoupling are becoming easier to study with the development of statistical inference methods for coupled systems, i.e., metapopulation dynamics.

@@

  \bi
  \item
    Particle filter (PF) methods are effective for inference on low-dimensional nonlinear partially observed stochastic dynamic systems. They scale exponentially badly.

\item Extending the successes of particle filter methods from time series data to metapopulation data is becoming possible.

\item Algorithms under consideration:\\
  {\bf
  Bagged filters (BF, IBF)\\
  Ensemble Kalman filter (EnKF, IEnKF)\\
  Guided intermediate resampling filter (GIRF, IGIRF)\\
  Block particle filter (BPF, IBPF)\\
  }

\item Filters estimate latent states and evaluate the likelihood.
\item Each filter has an iterated version which estimates parameters by repeated filtering using stochastic parameter perturbations.

\item These algorithms are all implemented in an R package, \texttt{spatPomp}.

@@

@@

\begin{frame}{He et al (2010) parameters; $r$ is rank correlation with $N_{1950}$}

\resizebox{12cm}{!}{
\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r}
\hline
  & $N_{1950}$ & $R_0$ & IP & LP & $\alpha$ & $a$ & $\iota$ & $\psi$ & $\rho$ & $\sigma_{SE}$\\
\hline
Halesworth & 2200 & 33.00 & 2.30 & 7.90 & 0.95 & 0.38 & 0.0091 & 0.64 & 0.75 & 0.075\\
\hline
Lees & 4200 & 30.00 & 2.10 & 8.50 & 0.97 & 0.15 & 0.0310 & 0.68 & 0.61 & 0.080\\
\hline
Mold & 6400 & 21.00 & 1.80 & 5.90 & 1.00 & 0.27 & 0.0140 & 2.90 & 0.13 & 0.054\\
\hline
Dalton in Furness & 11000 & 28.00 & 2.00 & 5.50 & 0.99 & 0.20 & 0.0390 & 0.82 & 0.46 & 0.078\\
\hline
Oswestry & 11000 & 53.00 & 2.70 & 10.00 & 1.00 & 0.34 & 0.0300 & 0.48 & 0.63 & 0.070\\
\hline
Northwich & 18000 & 30.00 & 3.00 & 8.50 & 0.95 & 0.42 & 0.0600 & 0.40 & 0.80 & 0.086\\
\hline
Bedwellty & 29000 & 25.00 & 3.00 & 6.80 & 0.94 & 0.16 & 0.0400 & 0.95 & 0.31 & 0.061\\
\hline
Consett & 39000 & 36.00 & 2.70 & 9.10 & 1.00 & 0.20 & 0.0730 & 0.41 & 0.65 & 0.071\\
\hline
Hastings & 66000 & 34.00 & 5.40 & 7.00 & 1.00 & 0.30 & 0.1900 & 0.40 & 0.70 & 0.096\\
\hline
Cardiff & 240000 & 34.00 & 3.10 & 9.90 & 1.00 & 0.22 & 0.1400 & 0.27 & 0.60 & 0.054\\
\hline
Bradford & 290000 & 32.00 & 3.40 & 8.50 & 0.99 & 0.24 & 0.2400 & 0.19 & 0.60 & 0.045\\
\hline
Hull & 300000 & 39.00 & 5.50 & 9.20 & 0.97 & 0.22 & 0.1400 & 0.26 & 0.58 & 0.064\\
\hline
Nottingham & 310000 & 23.00 & 3.70 & 5.70 & 0.98 & 0.16 & 0.1700 & 0.26 & 0.61 & 0.038\\
\hline
Bristol & 440000 & 27.00 & 4.90 & 6.20 & 1.00 & 0.20 & 0.4400 & 0.20 & 0.63 & 0.039\\
\hline
Leeds & 510000 & 48.00 & 11.00 & 9.50 & 1.00 & 0.27 & 1.2000 & 0.17 & 0.67 & 0.078\\
\hline
Sheffield & 520000 & 33.00 & 6.40 & 7.20 & 1.00 & 0.31 & 0.8500 & 0.18 & 0.65 & 0.043\\
\hline
Manchester & 700000 & 33.00 & 6.90 & 11.00 & 0.96 & 0.29 & 0.5900 & 0.16 & 0.55 & 0.055\\
\hline
Liverpool & 800000 & 48.00 & 9.80 & 7.90 & 0.98 & 0.30 & 0.2600 & 0.14 & 0.49 & 0.053\\
\hline
Birmingham & 1100000 & 43.00 & 12.00 & 8.50 & 1.00 & 0.43 & 0.3400 & 0.18 & 0.54 & 0.061\\
\hline
London & 3400000 & 57.00 & 13.00 & 13.00 & 0.98 & 0.55 & 2.9000 & 0.12 & 0.49 & 0.088\\
\hline
$\qquad r$ & 1 & 0.46 & 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
\hline
\end{tabular}
}

@@

  \begin{center}
\begin{tabular}{|l|rc|rc|rc|}
\hline
& Model~1& (p) & Model~2 & (p) & Model~3 & (p)
\\
\hline
Wheeler et al
& { -2731.3} & (15)
& -21957.3 & (6)
& { -17850.4} & (35)
\\
Lee et al
& -3030.9 & (20)
& -29367.4 & (6)
& -31840.8 & (29)
\\
Log-ARMA(2,1)
& -2803.7 & (4)  %% -0.5*5615.3+4
& { -18027.0} & (40)  %% -0.5*36133.9+40
& -18027.0 & (40)
\\
\hline
\end{tabular}
  \end{center}

@@

    Model 3 (stochastic, spatial): \\
    (We fit the model using IBPF, and beat the ARMA benchmark.

@@

@@

\nocite{bjornstad01,grenfell04,breto19,rebeschini15,ng02,ionides21,wheeler23,lee20,asfaw23arxiv}

>>>>

Joint work with
Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King

@@

  1--5 are largely solved, from a methodological perspective.\\
  6 is our immediate topic.\\
  7 is exciting but not the focus of this talk.

@@

\begin{frame}{Pre-vaccination weakly measles in England \& Wales}

@@

    Hence, \myemph{likelihood-based} methods.

      \item  For time series analysis, iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do this (\url{https://ionides.github.io/531w22/}). The science may be hard, but the statistics is becoming routine.
      \end{itemize}
  \end{frame}

@@

\item \myemph{Decoupling} leads to panel data analysis, by assumption.

  \item Iterated filtering methods extend to panel data (Breto et al, {\it Journal of the American Statistical Association}, 2019).

\item We must decide which parameters should be modeled as \myemph{shared} (i.e., the same for each unit) vs \myemph{unit-specific}.

%\item The consequences of decoupling are becoming easier to study with the development of statistical inference methods for coupled systems, i.e., metapopulation dynamics.

@@

  \bi
  \item
    Particle filter (PF) methods fail for high-dimensional systems. They scale exponentially badly.

  \item Algorithms with improved scalability include:\\

  {\bf
  Bagged filters (BF, IBF)\\
  Ensemble Kalman filter (EnKF, IEnKF)\\
  Guided intermediate resampling filter (GIRF, IGIRF)\\
  Block particle filter (BPF, IBPF)\\
  }

\item Filters estimate latent states and evaluate the likelihood.

  \item Each filter has an iterated version which estimates parameters by repeated filtering using stochastic parameter perturbations.

\item These algorithms are all implemented in the \texttt{spatPomp} R package.

@@

      \item IBPF has theoretical guarantees similar to BPF (Ning \& Ionides, {\it arXiv}, 2023).

@@

\begin{frame}{He et al (2010) estimates. $r$ is rank correlation with $N_{1950}$}

\resizebox{12cm}{!}{
\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r}
\hline
  & $N_{1950}$ & $R_0$ & IP & LP & $\alpha$ & $a$ & $\iota$ & $\psi$ & $\rho$ & $\sigma_{SE}$\\
\hline
Halesworth & 2200 & 33.00 & 2.30 & 7.90 & 0.95 & 0.38 & 0.0091 & 0.64 & 0.75 & 0.075\\
\hline
Lees & 4200 & 30.00 & 2.10 & 8.50 & 0.97 & 0.15 & 0.0310 & 0.68 & 0.61 & 0.080\\
\hline
Mold & 6400 & 21.00 & 1.80 & 5.90 & 1.00 & 0.27 & 0.0140 & 2.90 & 0.13 & 0.054\\
\hline
Dalton in Furness & 11000 & 28.00 & 2.00 & 5.50 & 0.99 & 0.20 & 0.0390 & 0.82 & 0.46 & 0.078\\
\hline
Oswestry & 11000 & 53.00 & 2.70 & 10.00 & 1.00 & 0.34 & 0.0300 & 0.48 & 0.63 & 0.070\\
\hline
Northwich & 18000 & 30.00 & 3.00 & 8.50 & 0.95 & 0.42 & 0.0600 & 0.40 & 0.80 & 0.086\\
\hline
Bedwellty & 29000 & 25.00 & 3.00 & 6.80 & 0.94 & 0.16 & 0.0400 & 0.95 & 0.31 & 0.061\\
\hline
Consett & 39000 & 36.00 & 2.70 & 9.10 & 1.00 & 0.20 & 0.0730 & 0.41 & 0.65 & 0.071\\
\hline
Hastings & 66000 & 34.00 & 5.40 & 7.00 & 1.00 & 0.30 & 0.1900 & 0.40 & 0.70 & 0.096\\
\hline
Cardiff & 240000 & 34.00 & 3.10 & 9.90 & 1.00 & 0.22 & 0.1400 & 0.27 & 0.60 & 0.054\\
\hline
Bradford & 290000 & 32.00 & 3.40 & 8.50 & 0.99 & 0.24 & 0.2400 & 0.19 & 0.60 & 0.045\\
\hline
Hull & 300000 & 39.00 & 5.50 & 9.20 & 0.97 & 0.22 & 0.1400 & 0.26 & 0.58 & 0.064\\
\hline
Nottingham & 310000 & 23.00 & 3.70 & 5.70 & 0.98 & 0.16 & 0.1700 & 0.26 & 0.61 & 0.038\\
\hline
Bristol & 440000 & 27.00 & 4.90 & 6.20 & 1.00 & 0.20 & 0.4400 & 0.20 & 0.63 & 0.039\\
\hline
Leeds & 510000 & 48.00 & 11.00 & 9.50 & 1.00 & 0.27 & 1.2000 & 0.17 & 0.67 & 0.078\\
\hline
Sheffield & 520000 & 33.00 & 6.40 & 7.20 & 1.00 & 0.31 & 0.8500 & 0.18 & 0.65 & 0.043\\
\hline
Manchester & 700000 & 33.00 & 6.90 & 11.00 & 0.96 & 0.29 & 0.5900 & 0.16 & 0.55 & 0.055\\
\hline
Liverpool & 800000 & 48.00 & 9.80 & 7.90 & 0.98 & 0.30 & 0.2600 & 0.14 & 0.49 & 0.053\\
\hline
Birmingham & 1100000 & 43.00 & 12.00 & 8.50 & 1.00 & 0.43 & 0.3400 & 0.18 & 0.54 & 0.061\\
\hline
London & 3400000 & 57.00 & 13.00 & 13.00 & 0.98 & 0.55 & 2.9000 & 0.12 & 0.49 & 0.088\\
\hline
$\qquad r$ & 1 & 0.46 & \cellcolor{yellow!50} 0.95 & 0.32 & 0.11 & 0.30 & 0.9300 & -0.93 & -0.20 & -0.330\\
\hline
\end{tabular}
}

@@

  \begin{center}
\begin{tabular}{|l|rc|rc|rc|}
\hline
& Model~1& (p) & Model~2 & (p) & Model~3 & (p)
\\
\hline
Wheeler et al
& \cellcolor{yellow!50} -2731.3 & \cellcolor{yellow!50}(15)
& -21957.3 & (6)
& \cellcolor{yellow!50} -17850.4 & \cellcolor{yellow!50}(35)
\\
Lee et al
& -3030.9 & (20)
& -29367.4 & (6)
& -31840.8 & (29)
\\
Log-ARMA(2,1) \hspace{-2.5mm}
& -2803.7 & (4)  %% -0.5*5615.3+4
& \cellcolor{yellow!50} { -18027.0} & \cellcolor{yellow!50} (40)  %% -0.5*36133.9+40
& -18027.0 & (40)
\\
\hline
\end{tabular}
  \end{center}

@@

    Model 3 (stochastic, spatial): \\
    We fit the model using IBPF, and beat the ARMA benchmark.

@@

\begin{frame}{Evidence for a decreasing transmission rate in Model 1}
\includegraphics[width=12cm]{haiti/Beta_trend_Figure-1.pdf}
\end{frame}

@@

\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23}
